---
title: "Adaptive gPCA Vignette"
author: "Julia Fukuyama"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adaptive gPCA Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
n---

# Adaptive gPCA Vignette

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(fig.width = 8, fig.height = 4)
```

Here we will describe how to use the adaptiveGPCA package. The package
implements the methods for structured dimensionality reduction
described in
[Fukuyama, J. (2017)](https://arxiv.org/abs/1702.00501). The general
idea is to obtain a low-dimensional representation of the data,
similar to that given by PCA, which incorporates side information
about the relationships between the variables. The output is similar
to a PCA biplot, but the variable loadings are regularized so that
similar variables are encouraged to have similar loadings on the
principal axes.

There are two main ways of using this package. The function
`adaptivegpca` will choose how much to regularize the variables
according to the similarities between them, while the function
`gpcaFullFamily` produces analogous output for a range of regularization
parameters. With this function, the results for the different
regularization parameters are inspected with the `visualizeFullFamily`
function, and the desired parameter is chosen manually. We will
illustrate both methods in this vignette. 

The data set we will use to illustrate the functionality of this
package is an antibiotic time course study[^1]. In this experiment, three subjects were
given two courses of antibiotics, and the abundances of bacteria in
the gut microbiome were measured before, during, and after each course
of antibiotics. The data is stored in a `phyloseq` object called
`AntibioticPhyloseq`, which contains variance-stabilized and centered
abundances on the `otu_table` slot, a phylogenetic tree describing the
relationships between the bacterial species measured in the `phy_tree`
slot, and information about the samples in the `sample_data` slot. We
want a low-dimensional representation of the samples which takes into
account the phylogenetic similarities between the bacteria, which is
what adaptive gPCA will provide for us. 


[^1]: Dethlefsen, L., and D. A. Relman. "Microbes and health sackler
colloquium: incomplete recovery and individualized responses of the
human distal gut microbiota to repeated antibiotic perturbation." Proc
Natl Acad Sci USA 108. Suppl 1 (2010): 4554-4561.

The first step is to load the required libraries and data. 
```{r}
library(adaptiveGPCA)
library(ggplot2)
library(phyloseq)
data(AntibioticPhyloseq)
theme_set(theme_bw())
```

Next, we create the inputs required for the `adaptivegpca`
function. If we have a `phyloseq` object, the function
`processPhyloseq` will create a centered data matrix (found in `pp$X`
below) and a similarity matrix based on the phylogeny (found in `pp$Q`
below), which can then be passed to the `adaptivegpca` function. If
you have another kind of data, you will need to create a centered data
matrix and a matrix describing the similarities between the variables.
```{r}
pp = processPhyloseq(AntibioticPhyloseq)
```

Next, we pass our data matrix and our similarity matrix to the
`adaptivegpca` function. The option `k = 2` means that we want a
two-dimensional representation of the data. 
```{r}
out.agpca = adaptivegpca(pp$X, pp$Q, k = 2)
```

Alternately, if we want to use the shiny interface to choose the
regularization parameter, we can first use the `gpcaFullFamily`
function to create a full set of ordinations and then use the
`visualizeFullFamily` function to visualize the biplots at each value of
the constraint.

`gpcaFullFamily` takes the same arguments as `adaptivegpca`. 

The `visualizeFullFamily` function is a shiny "gadget",
and so will open a browser window where you can visualize the data set
for a range of regularization parameters. Clicking "done" in this window will give as
output an object of the same format as that given by the `adaptivegpca`
function, the difference being that the value of `r` was chosen manually
instead of automatically. (The code below is only included as an
example and not evaluated in the vignette.) If you run this yourself
it will take little bit of time --- on my laptop it takes about a
minute and a half for the `gpcaFullFamily` function to run. The
`sample_data`/`sample_mapping` and `var_data`/`var_mapping` arguments
allow you to customize the visualization. Without these arguments, the
first two axes will be plotted for both the samples and the
variables. If you include these arguments, you can customize the plots
by providing an aesthetic mapping for ggplot to use. 
```{r, eval = FALSE}
out.ff = gpcaFullFamily(pp$X, pp$Q, k = 2)
out.agpca = visualizeFullFamily(out.ff,
                    sample_data = sample_data(AntibioticPhyloseq),
                    sample_mapping = aes(x = Axis1, y = Axis2, color = condition),
                    var_data = tax_table(AntibioticPhyloseq),
                    var_mapping = aes(x = Axis1, y = Axis2, color = Phylum))
```

In either case, we can plot the results. As desired, we get a nice
biplot representation where similar species are located in similar
positions. The sample scores on the principal axes are located in
`out.agpca$U` and the loadings of the variables on the principal axes
are located in `out.agpca$QV`. `out.agpca$r` gives the value of the
regularization parameter: values closer to 0 mean that there is only a
small amount of regularization, and 1 is the maximal amount of
regularization. 
```{r}
ggplot(data.frame(out.agpca$U, sample_data(AntibioticPhyloseq))) +
    geom_point(aes(x = Axis1, y = Axis2, color = type, shape = ind))
ggplot(data.frame(out.agpca$QV, tax_table(AntibioticPhyloseq))) +
    geom_point(aes(x = Axis1, y = Axis2, color = Phylum))
out.agpca$r
```

Finally, note that the `processPhyloseq` function also has an argument `ca`
(for correspondence analysis). This should be used with phyloseq
objects containing raw counts, and it will process a phyloseq object
so as to do an adaptive gPCA version of correspondence analysis (this
entails transforming counts to relative abunadnces, computing sample
weights based on the overall counts for the samples, and finally doing
a weighted centering of the relative abundances).
